### 2018-05-13 2_wmm.tex
complete some todos.
write up to Interpreter.

\begin{figure}[!htb]
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(0:EAX=1~/\textbackslash~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Possible candidate executions for the litmus test Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

@@

\section{The \cat{} language}
\label{ch:wmm:cat}

Weak memory models are defined via the \cat{} language~\cite{alglave2016syntax}.
This is a domain specific language for describing consistency properties of concurrent programs.
The \cat{} language combines expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, operations and assertions that are specific for operating with relations and executions.
%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.
In \cat{}, new relations can be defined via the keyword \texttt{let} and the following operators over relations~\cite{alglave2016syntax}. % 3.1.6 Relations between events

@@

\begin{enumerate}
  \item \textit{Unary operations}:
    \begin{itemize}
      \item \textit{the complement} of a relation \rel{r} is $\rel{\textasciitilde r}$,
      \item \textit{the transitive closure} of a relation \rel{r} is $\rel{r+}$,
      \item \textit{the reflexive closure} of a relation \rel{r} is $\rel{r?}$,
      \item \textit{the reflexive-transitive closure} of a relation \rel{r} is $\rel{r*}$, and
      \item \textit{the inverse} of a relation \rel{r} is $\invrel{r}$.
    \end{itemize}
  \item \textit{Binary operations:}
  \begin{itemize}
    \item \textit{the union} of two relations \rel{r1} and \rel{r2} is \rel{r1\,|\,r2},
    \item \textit{the intersection} of two relations \rel{r1} and \rel{r2} is \rel{r1\,\&\,r2},
    \item \textit{the difference} of two relations \rel{r1} and \rel{r2} is $\mathtt{r1\,\backslash\,r2}$, and
    \item \textit{the sequence} of two relations \rel{r1} and \rel{r2} is \rel{r1;r2}, which is defined as the set of pairs \rel{(x,y)} such that there exists an intervening \rel{z}, such that $\rel{(x,z)} \in \rel{r1}$ and $\rel{(z,y)} \in \rel{r2}$.
  \end{itemize}
\end{enumerate}

>>>>

\begin{figure}[!htb]
\centering
\begin{subfigure}[t]{.28\textwidth}
  \centering
  \includegraphics[width=1.2\linewidth]{img/my/sb-example/SB-dot.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub1}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-2.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub2}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-3.png}
  \caption{Final state: \texttt{(0:EAX=1,~1:EAX=1)}}
  \label{simple_wmm_x86_pic:sub3}
\end{subfigure}
\hfill
\begin{subfigure}[t]{.23\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{img/my/sb-example/SB-dot-4.png}
  \caption{Final state: \texttt{(0:EAX=0,~1:EAX=0)}}
  \label{simple_wmm_x86_pic:sub4}
\end{subfigure}
\hfill
\caption{Possible candidate executions for the litmus test Example~\ref{simple_wmm_x86}}
\label{simple_wmm_x86_pic}
\end{figure}

@@

\section{\cat{} language}
\label{ch:wmm:cat}

Weak memory models are defined via \cat{} language~\cite{alglave2016syntax}.
This is a domain specific language for describing consistency properties of concurrent programs.
The language combines expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, operations and assertions that are specific for operating with relations and executions.
%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.
In \cat{}, new relations can be defined via the keyword \texttt{let} and the following operators over relations~\cite{alglave2016syntax}. % 3.1.6 Relations between events

@@

\begin{enumerate}
  \item \textit{Unary operations}:
    \begin{itemize}
      \item \textit{the complement} of a relation \rel{r} is $\rel{\textasciitilde r}$,
      \item \textit{the transitive closure} of a relation \rel{r} is $\rel{r+}$,
      \item \textit{the reflexive closure} of a relation \rel{r} is $\rel{r?}$,
      \item \textit{the reflexive-transitive closure} of a relation \rel{r} is $\rel{r*}$, and
      \item \textit{the inverse} of a relation \rel{r} is $\invrel{r}$.
    \end{itemize}
  \item \textit{Binary operations:}
  \begin{itemize}
    \item \textit{the union} of two relations \rel{r1} and \rel{r2} is \rel{r1\,|\,r2},
    \item \textit{the intersection} of two relations \rel{r1} and \rel{r2} is \rel{r1\,\&\,r2},
      \item \textit{the difference} of two relations \rel{r1} and \rel{r2} is $\mathtt{r1 \relminus r2}$, and
    \item \textit{the sequence} of two relations \rel{r1} and \rel{r2} is \rel{r1;r2}, which is defined as the set of pairs \rel{(x,y)} such that there exists an intervening \rel{z}, such that $\rel{(x,z)} \in \rel{r1}$ and $\rel{(z,y)} \in \rel{r2}$.
  \end{itemize}
\end{enumerate}

### 2018-05-13 4_impl.tex
write up to Interpreter.

All recursive data structures are processed by units that implement the \textit{visitor} pattern~\cite{palsberg1998essence}.
This is a behavioural pattern that separates the program logic from the object implementation by specifying handling methods for each element of the object.
The visitor pattern performs the double-dispatching mechanism: the visiting class defines the method for accepting a visitor, and the visitor itself defines the 
To visit an instance \lstinline{x} by a visitor \lstinline{v}, the program should call the accept-method of \lstinline{x} and pass \lstinline{v} as an argument.
In our implementation, the visitor must carry out the continuation of traversing the recursive data structure itself.
The general structure of the visitor pattern is illustrated by the pseudo-java code in Figure~\ref{fig:visitor}.

@@

The input memory model language \cat{} is discussed in Section~\ref{ch:wmm:cat}.
The \tool{ANTLR} grammar for the \cat{} language was extracted from the parser used by \tool{herd} tool%
%
\footnote{\tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
%
written in OCaml.
%TODO: upload the grammar to github
%The input memory model 
%TODO: add reference to the grammar file in github repository OR include a short version as an appendix}

@@

%TODO: <not implemented yet>
The \wmodel{} is constructed by the stateless visitor \texttt{Cat2WmodelConverterVisitor} from the ANTLR syntax tree \texttt{CatParser.MainContext} of the memory-model defined in \cat{} language.
Currently, the visitor supports the small subset of the \cat{} language, which constitutes only non-functional declarative expressions of the language as the support for functional-style expressions requires implementing the full OCaml interpreter.
However, some most commonly used functional-style expressions may be supported by mapping them syntactically to corresponding W-elements by the \wmodel{} constructor.

Following the principle of transparency, the \wmodel{} constructor aborts its work with the \texttt{NotSupportedException} if received the unrecognised syntax construction (in contrast to the \porthos[1] approach in which the \texttt{null} value was produced in all exceptional states).

@@

%TODO: YOU REALLY NEED TO THINK MORE ABOUT TYPES ON Y, X, Z LEVELS + reconsider this paragraph (and others about types)
Currently, \porthos[2] handles only the primitive C types (such as \texttt{int}, \texttt{char}, \texttt{float}, etc.), which is modelled as an enumeration \texttt{XType}. %TODO: model it.
Array dimension is stored as an integer number at the Y-level (at the X- and Z- levels, \textbf{TODO: think about it}). %the array operations are converted to ).
In addition, \porthos[2] has a built-in extensible database storing the semantics of non-primitive C types stored by the \texttt{X2ZTypeConverter}, %TODO: Or XTypeHookManager?
which can be requested for converting an X-type to a~Z-type. %TODO: implement this.
%TODO: describe the XType (before) + implement this clearly! + definitely get rid of bit32

@@

\vspace{1em}
The following types of variables are detected as \textit{global}:
%TODO: not implemented
\begin{itemize}
\item a variable was declared as a pointer;
\item a variable whose address was accessed by any process;
\item a variable that was declared as a parameter of the process function;
%TODO: say sth about KERNEL ANALYSIS MODE BEFORE !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
\item a variable that was exported by the \texttt{extern} keyword (in the kernel-analysis mode, the functions \texttt{EXPORT\_SYMBOL} and \texttt{EXPORT\_SYMBOL\_GPL} also export symbols for dynamic linking); %https://lwn.net/Articles/154602/
\end{itemize}

@@

The X-compiler is the main component that transforms the recursive \ytree{} data structure to the plain \xgraph{} representation.

@@

The X-compiler is a complex processing unit, Figure~\ref{fig:compiler} illustrates the relationship between the principal components of the X-compiler in the UML language.

The X-interpreter is a stateful processing unit that carries the \xgraph{}-builder and provides \textit{action} methods for changing its state and thus the state of the builder.

<scheme of who calls whom>

@@

The invocation hooking module serves as a knowledge base that stores the semantics of functions.
%TODO: rename HookManager -> XHookManager
Once the X-compiler meets the function invocation, it calls the \texttt{XHookManager}, which tries to match the function signature (in case of program in C language -- only the function name) across all signatures it stores.
Once the signature matches, the hook manager intercepts the compilation process and interprets the method according to the algorithm it stores.
%TODO: say sth about the mode
The hook manager operates multiple invocation hooks depending on the user-defined mode.
All invocation hooks implement the interface \texttt{XInvocationHook}.
Current version of \porthos[2] contains two invocation hooks, the \texttt{} and \texttt{XKernelInvocationHook} that contains ...............................

%TODO: rename InvocationHook -> XInvocationHook
%, having access to the X-interpreter (see Section~\ref{ch:impl:proc:x-compiler:compilation}),

lambda-hook: knowledge base: InterceptionAction(BiFunction<XMemoryUnit, XMemoryUnit[], ? extends XEntity> action)

todo: kernel hooks
%TODO: kernel's : ACCESS_ONCE() access may be modeled as a volatile memory_order_relaxed access.  %http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2015/n4374.html

@@

>>>>

All recursive data structures are processed by units that implement the \textit{visitor} pattern~\cite{palsberg1998essence}.
This is a behavioural pattern that separates the program logic from the object implementation by specifying handling methods for each element of the object.
The visitor pattern performs the double-dispatching mechanism: the visiting class defines the method for accepting a visitor, and the visitor itself defines the 
<TODO!!!!>

To visit an instance \lstinline{x} by a visitor \lstinline{v}, the program should call the accept-method of \lstinline{x} and pass \lstinline{v} as an argument.
In our implementation, the visitor must carry out the continuation of traversing the recursive data structure itself.
The general structure of the visitor pattern is illustrated by the pseudo-java code in Figure~\ref{fig:visitor}.

@@

The input memory model language \cat{} is discussed in Section~\ref{ch:wmm:cat}.
The \tool{ANTLR} grammar for \cat{} language was extracted from the parser used by \tool{herd} tool%
%
\footnote{\tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
%
written in OCaml.
%TODO: upload the grammar to github
%The input memory model 
%TODO: add reference to the grammar file in github repository OR include a short version as an appendix}

@@

%TODO: <not implemented yet>
The \wmodel{} representation is constructed by the stateless visitor \texttt{Cat2WmodelConverterVisitor} from the ANTLR syntax tree \texttt{CatParser.MainContext} of the memory-model defined in \cat{} language.
Currently, the visitor supports the small subset of \cat{} language, which constitutes only non-functional declarative expressions of the language as the support for functional-style expressions requires implementing the full OCaml interpreter.
However, some most commonly used functional-style expressions may be supported by mapping them syntactically to corresponding W-elements by the \wmodel{} constructor.

Following the principle of transparency, the \wmodel{} constructor aborts its work with the \texttt{NotSupportedException} if met the unrecognised syntax construction (in contrast to the \porthos[1] approach in which the \texttt{null} value was produced in all exceptional states).

@@

%TODO: YOU REALLY NEED TO THINK MORE ABOUT TYPES ON Y, X, Z LEVELS + reconsider this paragraph (and others about types)
Currently, \porthos[2] handles only the primitive C types (such as \texttt{int}, \texttt{char}, \texttt{float}, etc.), which is modelled as an enumeration \texttt{XType}. %TODO: model it.
The array dimension is stored at the Y-level as an integer number (currently this information is not used as \porthos[2] does not support arrays, although this information can be used at the stage of pre-compilation for converting array elements to separate variables). %the array operations are converted to ).
In addition, \porthos[2] has a built-in extensible database storing the semantics of non-primitive C types stored by the \texttt{X2ZTypeConverter}, %TODO: Or XTypeHookManager?
which can be requested for converting an X-type to a~Z-type. %TODO: implement this.
%TODO: describe the XType (before) + implement this clearly! + definitely get rid of bit32

@@

%\vspace{1em}
The following types of variables are detected as \textit{global variables}:
%TODO: not implemented
\begin{itemize}
\item a variable was declared as a pointer;
\item a variable whose address was accessed by any process;
\item a variable that was declared as a parameter of the process function;
%TODO: say sth about KERNEL ANALYSIS MODE BEFORE !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
\item a variable that was exported by the \texttt{extern} keyword (in the kernel-analysis mode, the functions \texttt{EXPORT\_SYMBOL} and \texttt{EXPORT\_SYMBOL\_GPL} also export symbols for dynamic linking); %https://lwn.net/Articles/154602/
\end{itemize}

@@

The X-compiler is the main component that transforms the recursive \ytree{} data structure to the plain \xgraph{} representation.
It is a complex processing unit; Figure~\ref{fig:compiler} illustrates the relationship between the principal components of the X-compiler in the UML language.

<TODO: in picture change: xmemorymaager belongs to the program interpreter!>

@@

The main class representing the X-compiler is \texttt{Y2XConverter}. %TODO: draw it on Figure.
It receives as input the \ytree{}, the memory model kind and user settings (for instance, the interpreter mode defining the set of invocation hooks enabled during the analysis run).
The \texttt{Y2XConverter} creates an instance of the stateless visitor \texttt{Y2XConverterVisitor} that traverses the \ytree{} while invoking the stateful interpreter \texttt{XInterpreter}.
The \texttt{XInterpreter} carries the \xgraph{}-builder and provides \textit{action} methods for changing its state and thus the filling in the builder.
The interpreter requests the managers such as \texttt{XMemoryManager}, \texttt{XHookManager}, \texttt{XTypeManager} for additional information.

<TODO: ADD XTypeManager to the picture!>

We distinguish three types of processes (each implement the \texttt{XInterpreter} interface):
\begin{itemize}[noitemsep,topsep=0pt]
\item the prelude process \texttt{XPreludeInterpreter} (to be executed before all other processes) that allows only declaration of local or shared variables, computations and memory operations;
\item the postlude process \texttt{XPostludeInterpreter} (to be executed after all other processes) that allows only declaration of local variables, memory operations, computations and assertions;
\item the process \texttt{XProcessInterpreter} (to be executed in parallel) that allows all X-compiler interface operations except declaration of shared variables %(that should be declared during pre-compilation stage or during compilation of the prelude process)
and program assertions (declarations of local variables, computations, barriers, unconditional jumps, non-linear branching statements, method calls).
\end{itemize}

As the prelude process can declare shared variables, its compilation must go first.
%TODO: think about sorting processes at the pre-compilation stage
For transparency reasons, the first event of a process must be the \texttt{XEntryEvent} and the last events must be events of type \texttt{XExitEvent}.

@@

The invocation hooking module serves as a knowledge base that stores the semantics of functions.
%TODO: rename HookManager -> XHookManager
Once the X-compiler meets the function invocation, it calls the \texttt{XHookManager}, which tries to match the function signature (in case of program in C language -- only the function name) across all signatures it stores.
If the signature matches, the hook manager intercepts the function call and interprets the hook action instead of invoking the function directly.
%TODO: say sth about the mode
The hook manager operates multiple invocation hooks depending on the user-defined mode.
All invocation hooks implement the interface \texttt{XInvocationHook}.
The result of an invocation hook interception is the \texttt{XInvocationHookAction}, a delayed operation implemented on the top of lambda functions of Java.
The hook action is invoked with actual arguments and returns an arbitrary \texttt{XEntity} as the result of invocation.

Current version of \porthos[2] contains two invocation hooks, the \texttt{XPorthosInvocationHook} for describing the compatibility-mode functions of the input \porthos[2] input language, and \texttt{XKernelInvocationHook} for describing the Linux kernel-specific functions.
An invocation hook can model any type of operations (memory, fence, computational, etc.).
For instance, the \texttt{XPorthosInvocationHook} intercepts calls \texttt{`location.store(atomic, register)'} and processes the store operation with respect to the specified atomic.
%TODO: more about described functions
%TODO: rename InvocationHook -> XInvocationHook

%TODO: kernel's : ACCESS_ONCE() access may be modeled as a volatile memory_order_relaxed access.  %http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2015/n4374.html

@@

contexts

computation: how not to collect (in which case we loose)
